---
title: "Computing OLS and Inference, CIs, Hypothesis Testing, Plots"
author: "PSTAT126"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    # number_sections: yes
    toc_depth: 3
  fig_caption: yes
  highlight: haddock
  # number_sections: no
  df_print: paged
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
fontfamily: mathpazo
editor_options: null
chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)
```

```{r, include=FALSE}
rm(list=ls())
library(tidyverse) # Easily Install and Load the 'Tidyverse'
library(palmerpenguins) # Palmer Archipelago (Antarctica) Penguin Data
library(ggplot2)
library(skimr)
```

## Computing OLS estimators in simple linear regression

Dataset: Adelie and Gentoo Penguins

-   **Question: Can we predict body mass in grams by a penguins bill length in mm?**
-   What is the predictor?
-   What is the response variable?

```{r, fig.align='center', fig.height=4}
data("penguins")

# Plot density of body mass
penguins %>% 
  ggplot(aes(x = body_mass_g)) + 
  geom_density()+
  theme_minimal()

# plot density of bill_length
penguins %>% 
  ggplot(aes(x = bill_length_mm)) + 
  geom_density()+
  theme_minimal()

# Overlay by species
penguins %>% 
  ggplot(aes(x = bill_length_mm, fill = species)) + 
  geom_density(alpha = 0.5)+
  scale_fill_viridis_d() +  
  theme_minimal()

# Drop one species
penguins_noChinstrap <- penguins %>% 
  filter(species != "Chinstrap") %>% 
  drop_na(bill_length_mm, body_mass_g)

# skim(penguins_noChinstrap)

# plot of data
ggplot(data = penguins_noChinstrap,
       aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point(color = "slategray3", alpha = 0.9, size = 1.75) +
  labs(title = "Body Mass vs Bill Length",
       x = "Bill Length (mm)",
       y = "Body Mass (g)") +
  # Note: to modify anything about the plot using theme, need to call it after
  # theme_minimal()
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))  


```

## The lm() function

### Fit model

```{r}
model <- lm(body_mass_g ~ bill_length_mm , data = penguins_noChinstrap)
```

```{r}
summary(model)
```

```{r}
coef(model) # Estimates for b0 and b1
model$coefficients
```

When bill length increases by 1mm, body mass increases by 141 grams.

```{r}
# estimate for the standard deviation of the random errors
summary(model)$sigma 
```

On average, the observed response values differ from the fitted regression line by about 402.5 grams.

```{r, fig.align='center', fig.height=4}

par(mfrow = c(1, 1))
resid_model <- tibble(residuals = residuals(model))

ggplot(data = resid_model,
       aes(x = residuals)) +
  geom_histogram(color = "white",
            fill = "seagreen",
            bins = 15) +
  labs(x = "Residuals",
       title = "Histogram of Residuals") +
  theme_minimal() 
```

```{r}
# Check if sum of residuals add up to 0 
round(sum(model$residuals))
```

### Point estimate

Now that we have a linear model, we can predict body mass when we have a new data point for bill length.
Note that we cannot input just any value of bill length, it has to be within the range of the existing value.

```{r}
summary(penguins_noChinstrap$bill_length_mm)
predict(model, newdata = data.frame(bill_length_mm = 45))
```



\newpage

## Review of Normal, Chi-squared and t-distribution

### Normal Distribution

Let $X_1, X_2,...,X_n \stackrel{iid}{\sim} N(\mu, \sigma^2)$.
Then the pdf is given for any $\mu\in \mathbb{R},\,x\in \mathbb{R},\,\sigma>0$:

$$f_X(x) = \frac{1}{\sqrt{2 \pi \sigma^2}} \exp{\frac{-(x-\mu)^2}{2\sigma^2}}$$

Then define the following: $E(X_i)=\mu, Var(X_i)=\sigma^2\,\forall i$ The CDF is given by, $F_X(x)=P(X \le x)= 1-P(X > x)$ For $\mu=0, \sigma=1$, we have the standard normal $Z \sim N(0,1)$:

Some properties are as follows:

1)  $P(Z< z)= P(Z> -z)$
2)  $P(|Z| > z)= 2P(Z>z)= 2P(Z<-z)$
3)  if $P(Z> z_{\alpha})=\alpha$, then

$$\begin{aligned}
P(z_{1-\alpha/2}< Z < z_{\alpha/2})&=P(-z_{\alpha/2}< Z < z_{\alpha/2})\\
&=1-\alpha
\end{aligned}$$

```{r}
set.seed(5)
# consider the x-values
z <- rnorm(1000, 0,1)
x1 <- rnorm(1000, 0,.5)
x2 <- rnorm(1000, 0, .75)
x3 <- rnorm(1000, -2, 1.5)


df <- data.frame(
  value = c(z, x1, x2, x3),
  variable = factor(rep(c("z ~ N(0,1)",
                          "x1 ~ N(0,0.5)",
                          "x2 ~ N(0,0.75)",
                          "x3 ~ N(-2,1.5)"),
                        each = 1000))
)

# plot densities
ggplot(df, aes(x = value, color = variable)) +
  geom_density(size = .7) +
  labs(title = "Density Comparison",
       x = "Value",
       y = "Density",
       color = "Distribution") +
  theme_minimal()
```

### Chi-Squared Distribution

If $X_1, \dots, X_n \stackrel{iid}{\sim} N(0,1)$, then $$C = \sum_{i=1}^n X_i^2 \sim \chi_n^2$$ with $E(C) = n$ and $Var(C) = 2n$.

```{r}
x_chi <- seq(0, 20, length.out = 100)
chi1 <- dchisq(x_chi, df = 3)
chi2 <- dchisq(x_chi, df = 5)
chi3 <- dchisq(x_chi, df = 10)

plot(x_chi, chi1, type = "l", col = 1, lwd = 2, ylab = "Density", 
     main = "Chi-Squared Distributions")
lines(x_chi, chi2, col = 2, lwd = 2)
lines(x_chi, chi3, col = 3, lwd = 2)
legend("topright", legend = c("df = 3", "df = 5", "df = 10"), col = 1:3, lty = 1, lwd = 2)
```

### t-Distribution

Given $C \sim \chi_n^2$ and $Z \sim N(0,1)$ (independent), the t-statistic is: $$t = \frac{Z}{\sqrt{C/n}} \sim t_n.$$

### Comparison with Normal Distribution

```{r}
curve(dnorm(x), from = -4, to = 4, col = "blue", lwd = 2, 
      ylab = "Density", main = "t-Distribution vs Normal")
curve(dt(x, df = 3), from = -4, to = 4, col = "red", lwd = 2, add = TRUE)
curve(dt(x, df = 10), from = -4, to = 4, col = "green", lwd = 2, add = TRUE)
legend("topright", legend = c("Normal", "t (df=3)", "t (df=10)"), 
       col = c("blue", "red", "green"), lwd = 2)
```

\newpage

## Confidence Intervals for intercept and slope estimates

Can calculate a 90% confidence interval by entering values into formula:

-   **Intercept**

$$\hat\beta_0 \pm (t_{(1-\alpha/2), n-2}\boldsymbol{SE}(\hat\beta_0))$$

-   **Slope**

$$\hat\beta_1 \pm (t_{(1-\alpha/2), n-2}\boldsymbol{SE}(\hat\beta_1))$$

```{r}
b0 <- model$coefficients[[1]]
b1 <- model$coefficients[[2]] 
y_hat <- b0+ b1* penguins_noChinstrap$bill_length_mm
y <- penguins_noChinstrap$body_mass_g
x <-  penguins_noChinstrap$bill_length_mm
x_bar <- mean(x)
e <- y - y_hat

n <- length(x)
sigma_2_hat <- sum(e^2) / (n-2)
sigma_hat <- sqrt(sigma_2_hat) 
Sxx <- sum((x - x_bar)^2)

se_b0 <- sqrt(sigma_2_hat*(1/n + (x_bar^2)/Sxx)) # se of intercept
se_b1 <-  sqrt(sigma_2_hat/Sxx) # se of slope

t_pct <- qt(p = 0.95, df = n - 2) # t-statistic
```

```{r}
CI_b0_90 <-  c(b0 - t_pct*se_b0, b0 + t_pct*se_b0) # 90% CI for b0
CI_b1_90 <-  c(b1 - t_pct*se_b1, b1 + t_pct*se_b1) # 90% CI for b1
CI_b0_90
CI_b1_90
```

Can also use the `confint` function

```{r}
#?confint
confint(model, level = 0.95) # 95% CI

confint(model, level = 0.90) # 90% CI

```

## Hypothesis Testing

**Hypothesis testing of** $\beta_0, \beta_1$

Want to test:

$H_0:\beta_0 = 0$ vs. $H_1:\beta_0 \neq 0$\
$H_0:\beta_1 = 0$ vs. $H_1:\beta_1 \neq 0$\
Let $\alpha = 0.05$

```{r}
t_b0 <- (b0-0)/se_b0
t_b1 <- (b1-0)/se_b1
t_b0
t_b1
```

-   For distributions in R, p stands for "probability", the cumulative distribution function (c. d. f.).

```{r}
p0 <- 2*(1 - pt(abs(t_b0), df = n-2))
p1 <- 2*(1 - pt(abs(t_b1), df = n-2))

alpha <- 0.05
p0 > alpha
p1 > alpha

p0
p1
```

Reject Null Hypothesis for both $\beta_0, \beta_1$

We can also conduct hypothesis test using the regression output

```{r}
summary(model)
```

p-values for both $\beta_0$ and $\beta_1$ are small (1.65e-15 and \< 2e-16 respectively), we reject the null hypothesis.
